Experiences with Planning for Natural Language Generation

### 2008-06-23 trunk/Papers/experiences/domains.tex
small textual changes

Consider a knowledge base containing the individuals $r_1$ and $r_2$, and a
set of attributes encoding the fact that $r_1$ and $r_2$ are rabbits, $r_1$
is white and $r_2$ is brown, and $r_1$ sleeps.  Say that we want to express
the information $\{\mathsf{sleep}(r_1)\}$ using the tree-adjoining grammar
shown in Fig.~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists
of \emph{elementary trees} (i.e., the disjoint trees in the figure), each
of which contributes certain \emph{semantic content}. We can instantiate
these trees by substituting individuals for \emph{semantic roles}, such as
$\mathsf{self}$ and $\mathsf{subj}$, and then combine the tree instances as
shown in Fig.~\ref{fig:white-rabbit-sleeps-deriv} to obtain the sentence
``The white rabbit sleeps''. The SPUD algorithm computes this grammatical
derivation by starting with the elementary tree for ``sleeps''. This tree
satisfies the need to convey the semantic information, but introduces a
need to generate a noun phrase (NP) for the subject, which also refers
uniquely to the target referent $r_1$. In a second step, SPUD substitutes
the tree for ``the rabbit'' into the open NP leaf, which makes the
derivation grammatically complete. Since there are two different
individuals that could be described as ``the rabbit''---technically, $r_2$
is still a \emph{distractor} (i.e., a \todo{one line definition here}) for
the referring expression---we are still not finished. To complete the
derivation, the tree for ``white'' is added to the existing structure by
the \emph{adjunction} operation, making the derivation syntactically and
semantically complete.

@@

The process described above has clear parallels to planning: we manipulate
a state by applying actions in order to achieve a goal. We can make this
connection even more precise by translating the SPUD problem into a
planning problem. For instance, Fig.~\ref{fig:white-rabbit-as-planning}
shows the corresponding PDDL actions for the above generation task, where
each action corresponds to an operation that adds a single elementary tree
to the derivation.  In each case, the first parameter of the action is a
node name in the derivation tree, and the remaining parameters stand for
the individuals to which the semantic roles will be instantiated.  The
syntactic preconditions and effects are encoded using $\mathsf{open}$ and
$\mathsf{canadjoin}$ literals; the status of each referring expression is
tracked using $\mathsf{distractor}$ literals. Notice that the action
effects contain terms of the form $\mathsf{subj}(u)$, which construct new
node names. In order to meet the syntactic requirements of PDDL, these
terms can be eliminated by estimating an upper bound $n$ for the plan
length, making $n$ copies of each action, ensuring that copy $i$ can only
be applied in step $i$, and replacing the term $\mathsf{subj}(u)$ in an
action copy by the constant $\mathsf{subj}_i$.

Once we've defined appropriate actions, we can solve the generation
problem as an ordinary planning problem. For instance, the following plan
solves our previous example:
%
\begin{enumerate}
\item $\mathsf{sleeps}(\mathsf{root}, r_1)$
\item $\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}), r_1)$
\item $\mathsf{white}(\mathsf{subj}(\mathsf{root}), r_1)$
\end{enumerate}
%
Using this plan, the grammatical derivation in
Fig.~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the generated
sentence, can be systematically reconstructed. Thus, we can solve the
sentence generation problem via the detour through planning and bring
current search heuristics for planning to bear on generation.

@@

A map of an example GIVE world is shown in
Fig.~\ref{fig:give-development-world}.  In this world, the user's task is
to pick up a trophy in the top left room.  The trophy is hidden in a safe
behind a picture, so the user must first move the picture out of the way
and open the safe by pushing a sequence of buttons (the small square boxes
on the walls) in the correct order.  In order to get to all these buttons,
the user must also open the lower door and deactivate the alarm tile by
pushing further buttons.  To simplify both the planning and the NLG task,
the world is discretized into tiles of equal size; the user can turn by 90
degree steps in either direction, and can move from the centre of one tile
to the centre of the next. Some of the actions the user can take are shown
in Fig.~\ref{fig:give-planning}, in PDDL syntax.

>>>>

Consider a knowledge base containing the individuals $r_1$ and $r_2$, and a
set of attributes encoding the fact that $r_1$ and $r_2$ are rabbits, $r_1$
is white and $r_2$ is brown, and $r_1$ sleeps.  Say that we want to express
the information $\{\mathsf{sleep}(r_1)\}$ using the tree-adjoining grammar
shown in Figure~\ref{fig:white-rabbit-sleeps-grammar}. This grammar consists
of \emph{elementary trees} (i.e., the disjoint trees in the figure), each
of which contributes certain \emph{semantic content}. We can instantiate
these trees by substituting individuals for \emph{semantic roles}, such as
$\mathsf{self}$ and $\mathsf{subj}$, and then combine the tree instances as
shown in Figure~\ref{fig:white-rabbit-sleeps-deriv} to obtain the sentence
``The white rabbit sleeps''. The SPUD algorithm computes this grammatical
derivation by starting with the elementary tree for ``sleeps''. This tree
satisfies the need to convey the semantic information, but introduces a
need to generate a noun phrase (NP) for the subject, which also refers
uniquely to the target referent $r_1$. In a second step, SPUD substitutes
the tree for ``the rabbit'' into the open NP leaf, which makes the
derivation grammatically complete. Since there are two different
individuals that could be described as ``the rabbit''---technically, $r_2$
is still a \emph{distractor} (i.e., a \todo{one line definition here}) for
the referring expression---we are still not finished. To complete the
derivation, the tree for ``white'' is added to the existing structure by
the \emph{adjunction} operation, making the derivation syntactically and
semantically complete.

@@

The process described above has clear parallels to planning: we manipulate
a state by applying actions in order to achieve a goal. We can make this
connection even more precise by translating the SPUD problem into a
planning problem. For instance, Figure~\ref{fig:white-rabbit-as-planning}
shows the corresponding PDDL actions for the above generation task, where
each action corresponds to an operation that adds a single elementary tree
to the derivation.  In each case, the first parameter of the action is a
node name in the derivation tree, and the remaining parameters stand for
the individuals to which the semantic roles will be instantiated.  The
syntactic preconditions and effects are encoded using $\mathsf{open}$ and
$\mathsf{canadjoin}$ literals; the status of each referring expression is
tracked using $\mathsf{distractor}$ literals. Notice that the action
effects contain terms of the form $\mathsf{subj}(u)$, which construct new
node names. In order to meet the syntactic requirements of PDDL, these
terms can be eliminated by estimating an upper bound $n$ for the plan
length, making $n$ copies of each action, ensuring that copy $i$ can only
be applied in step $i$, and replacing the term $\mathsf{subj}(u)$ in an
action copy by the constant $\mathsf{subj}_i$.

Once we've defined appropriate actions, we can solve the generation
problem as an ordinary planning problem. For instance, the following plan
solves our previous example:
%
\begin{enumerate}
\item $\mathsf{sleeps}(\mathsf{root}, r_1)$
\item $\mathsf{rabbit}(\mathsf{subj}(\mathsf{root}), r_1)$
\item $\mathsf{white}(\mathsf{subj}(\mathsf{root}), r_1)$
\end{enumerate}
%
Using this plan, the grammatical derivation in
Figure~\ref{fig:white-rabbit-sleeps-deriv}, and therefore the generated
sentence, can be systematically reconstructed. Thus, we can solve the
sentence generation problem via the detour through planning and bring
current search heuristics for planning to bear on generation.

@@

A map of an example GIVE world is shown in
Figure~\ref{fig:give-development-world}.  In this world, the user's task is
to pick up a trophy in the top left room.  The trophy is hidden in a safe
behind a picture, so the user must first move the picture out of the way
and open the safe by pushing a sequence of buttons (the small square boxes
on the walls) in the correct order.  In order to get to all these buttons,
the user must also open the lower door and deactivate the alarm tile by
pushing further buttons.  To simplify both the planning and the NLG task,
the world is discretized into tiles of equal size; the user can turn by 90
degree steps in either direction, and can move from the centre of one tile
to the centre of the next. Figure~\ref{fig:give-planning} shows some of the
available actions in PDDL syntax.

### 2008-06-23 trunk/Papers/experiences/experiments.tex
small textual changes

The results of this experiment are shown in
Fig.~\ref{fig:runtimes-crisp}. The input parameters $(m,n)$ are plotted
(in lexicographic order) on the horizontal axis and the runtime is shown in
milliseconds on the vertical axis. These results reveal a number of
interesting insights. First, FF significantly outperforms
SGPLAN in this domain.\footnote{Experiments with SGPLAN use a
 pre-release version of SGPLAN, kindly provided by Chih-Wei Hsu. The
 release version of SGPLAN 5.2.2 had a bug, causing it to crash on some
 instances.}
Second, FF's runtime is dominated by its initial grounding step, in which
it computes the ground instances of all predicates and actions used in the
planning problem to avoid unnecessary instantiations during search. In
particular, the ratio of grounding time to total runtime is generally above
85\%, and rises to above 99\% at $m=11$, which is still a small universe in
this application.\footnote{The
  ``grounding'' time reported here is what FF reports as ``time spent:
  instantiating action templates''.}  This is also reflected in the
much higher sensitivity of FF to the domain size: if we fix $n=1$, FF takes
60 ms to compute a plan at $m=1$, but 2.4 seconds at $m=10$. By contrast,
our naive GraphPlan implementation takes 30 ms at $m=1$ and still only 50
ms at $m=10$.

@@

To evaluate the performance of the planners on the planning problems
that arise in the GIVE domain, we construct a sequence of test worlds
as illustrated in Fig.~\ref{fig:give-minimal}.  These worlds consist
of a $2n$ by $h$ grid of positions, such that there are buttons at
positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i \leq n$.  The player
starts in position $(1,1)$ and must press all buttons in order to
complete the game.  The world is generated as a GIVE world
description, and then converted into a planning problem by the GIVE
software. \todo{at what level of detail should we describe this?}

@@

Some results from this experiment (for $h=20$, with $n$ ranging from
$1$ to $40$) are shown in Fig.~\ref{fig:give-runtime-minimal}.  The
most obvious result is that FF is unable to solve any problems beyond
$n=13$ on the experimentation machine within the memory limit of 1 GB.
SGPLAN, on the other hand, solves instances beyond $n=40$ without
major problems.  The time spent on grounding is not a major factor in
either planner, probably because the planners need more time to
actually compute the plan -- at $n=40$, the optimal plan has a length
of about 1600!

@@

Finally, we varied the GIVE world in order to judge the effect of the
universe size in the GIVE domain (see Fig.~\ref{fig:give-junk}).  We
added another $w$ by $h$ positions to the right of a minimal world as
in Experiment 2.  Other than that, we left the initial state and goal
untouched, i.e.\ the new positions were not needed in any plan; this
approximates the situation in the actual GIVE domain, where most grid
positions never need to be used.  As before, we generate a GIVE world
description and then convert it into a planning problem.

@@

Some results for $h=20$ and $n=5$ (with $w$ ranging from $1$ to $70$)
are shown in Fig.~\ref{fig:give-runtime-junk}.  Again, FF runs out of
memory, this time at $w=17$, whereas SGPLAN happily solved inputs
beyond $w=70$.  However, unlike in Experiment 2, both planners now
spend a substantial propertion of their time on grounding again.  In
SGPLAN, this translates to a ``parsing time'' (which
\todo{presumably} includes grounding) which grows from 180 ms to 21.7
seconds as $w$ grows from $1$ to $75$; the rest of the runtime (which
includes the search) only grows from 400 ms to 2.3 seconds.  This
difference is particularly dramatic given that the actual optimal plan
is an identical plan of about 100 steps for each $w$.

>>>>

The results of this experiment are shown in
Figure~\ref{fig:runtimes-crisp}. The input parameters $(m,n)$ are plotted
(in lexicographic order) on the horizontal axis and the runtime is shown in
milliseconds on the vertical axis. These results reveal a number of
interesting insights. First, FF significantly outperforms
SGPLAN in this domain.\footnote{Experiments with SGPLAN use a
 pre-release version of SGPLAN, kindly provided by Chih-Wei Hsu. The
 release version of SGPLAN 5.2.2 had a bug, causing it to crash on some
 instances.}
Second, FF's runtime is dominated by its initial grounding step, in which
it computes the ground instances of all predicates and actions used in the
planning problem to avoid unnecessary instantiations during search. In
particular, the ratio of grounding time to total runtime is generally above
85\%, and rises to above 99\% at $m=11$, which is still a small universe in
this application.\footnote{The
  ``grounding'' time reported here is what FF reports as ``time spent:
  instantiating action templates''.}  This is also reflected in the
much higher sensitivity of FF to the domain size: if we fix $n=1$, FF takes
60 ms to compute a plan at $m=1$, but 2.4 seconds at $m=10$. By contrast,
our naive GraphPlan implementation takes 30 ms at $m=1$ and still only 50
ms at $m=10$.

@@

In the second experiment, we evaluate the performance of the planners on
problems arising in the GIVE domain. We construct a series of test worlds,
similar to the one illustrated in Figure~\ref{fig:give-minimal}. These worlds
consist of a $2n$ by $h$ grid of positions, such that there are buttons at
positions $(2i-1,1)$ and $(2i,h)$ for $1 \leq i \leq n$. The player starts
in position $(1,1)$ and must press all buttons in order to complete the
game. The world is generated as a GIVE world description, and then
automatically converted into a planning problem by the GIVE software.  (We
refer the reader to \todo{cite?} for more details about this process.)

@@

Some results from this experiment (for $h=20$, with $n$ ranging from
$1$ to $40$) are shown in Figure~\ref{fig:give-runtime-minimal}.  The
most obvious result is that FF is unable to solve any problems beyond
$n=13$ on the experimentation machine within the memory limit of 1 GB.
SGPLAN, on the other hand, solves instances beyond $n=40$ without
major problems.  The time spent on grounding is not a major factor in
either planner, probably because the planners need more time to
actually compute the plan -- at $n=40$, the optimal plan has a length
of about 1600!

@@

Finally, we varied the GIVE world in order to judge the effect of the
universe size in the GIVE domain (see Figure~\ref{fig:give-junk}).  We
added another $w$ by $h$ positions to the right of a minimal world as
in Experiment 2.  Other than that, we left the initial state and goal
untouched, i.e.\ the new positions were not needed in any plan; this
approximates the situation in the actual GIVE domain, where most grid
positions never need to be used.  As before, we generate a GIVE world
description and then convert it into a planning problem.

@@

Some results for $h=20$ and $n=5$ (with $w$ ranging from $1$ to $70$)
are shown in Figure~\ref{fig:give-runtime-junk}.  Again, FF runs out of
memory, this time at $w=17$, whereas SGPLAN happily solved inputs
beyond $w=70$.  However, unlike in Experiment 2, both planners now
spend a substantial propertion of their time on grounding again.  In
SGPLAN, this translates to a ``parsing time'' (which
\todo{presumably} includes grounding) which grows from 180 ms to 21.7
seconds as $w$ grows from $1$ to $75$; the rest of the runtime (which
includes the search) only grows from 400 ms to 2.3 seconds.  This
difference is particularly dramatic given that the actual optimal plan
is an identical plan of about 100 steps for each $w$.
